### 2025-07-02 manuscript/part2.tex
before travel

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as 
  the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
  from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
  All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
  which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textbf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
  dependency injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
  tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
  state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
  Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possible. Moreover, Langium enables developers to swap in
  custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec}.
  By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
  However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

The generated module is created automatically from the grammar definition and provides two DI fragments: the GeneratedSharedModule, which registers shared-core services such as AST reflection, and the GeneratedModule, which configures the parser, Chevrotain grammar factory, and language metadata. Conversely, the main module is handwritten: it imports Langium’s default core and shared modules, injects the generated fragments, declares custom services (validators, scope providers, caches), invokes the \texttt{inject} function to assemble the complete service container, and exposes a \texttt{createServices} factory.

Although the core logic resides in `/src`, files of interest at the project root include the standard npm `package.json` and two Langium-specific configs:
\code{language-configuration.json}, which governs editor behaviors such as comment delimiters, bracket pairs and auto-closing rules, and
\code{langium-config.json}, which declares settings like the language ID, grammar entry point, target file extensions and maximum lookahead.
In particular, the `syntaxes` entry drives the generation of a `.monarch.ts` module (for Monaco-based editors) and a `.tmLanguage.json` file (for TextMate-based engines).
The TextMate grammar maps regex patterns to semantic scopes that underpin VS Code's coloring, bracket
matching and code folding, and is the one applied for the extension. Customizing these definition files allows developers to implement nuanced, context-sensitive highlighting—such as semantic theming
and advanced pattern recognition—that cannot be exhaustively defined by grammar rules alone \cite{EDKarlsson2022Discussion604}.

@@

\section{Language Server Protocol}
control flow of lsp from file to lsp to editor

>>>>

\begin{description}
  \item[\code{language/}] Assembles the core of any Langium project. It includes the \code{.langium} file, in which the custom language grammar is defined, as well as 
  the code that starts the Language Server with all necessary services bundled.
  \item[\code{cli/}] Provides functionality for CLI interactions, command parsing, and generators. The generators in this folder invoke processes that generate artifacts
  from the language and input files; they are typically used only offline and are called once rather than continuously via the CLI and support efficient pipelining.
  \item[\code{extension/}] Contains the VS Code extension bootstrap code needed to connect to the Language Server and utilize LSP support features.
  All mentioned folders except the one containing CLI logic are bundled into the extension itself.
  \item[\code{generated/}] Contains artifacts produced from the grammar definition in the \code{.langium} file. This includes TypeScript AST type definitions,
  which are essential for further analysis and validation, as well as the preliminary parsed grammar rules formatted for the Chevrotain engine to generate parsers.
  \item[\textsf{Services}] The Services in Langium are a collection of utility classes and helper functions offered to every component of a language project via
  dependency injection. They cover an extensive range of functionality, from LSP-specific features - such as providers for context-aware code completion, rich hover
  tooltips, signature help, go-to-definition and find-references - to core language utilities that expose parsed documents as both AST and CST models, manage workspace
  state and context information (including indexing, scoping, and name resolution), synchronize the in-memory document model and perform validation checks.
  Because the parts are wired up as an injectable service, precise consumptions of the capabilities needed is possible. Moreover, Langium enables developers to swap in
  custom implementations of services by extending the corresponding default classes, and these services have to be wired into the DI container manually \ref{sec}.
  By default, the validation logic is located in the \code{language/} directory; subsequently, I placed additional custom services there.
  However, as these services grow in both number and complexity, it may be more appropriate to organize them in a dedicated module or folder.
\end{description}

Both module files share analogous responsibilities by encapsulating dependency-injection setup and service registration for the language environment.
The generated module is produced automatically from the grammar definition and offers two DI fragments, one that registers shared-core services such as AST
reflection and another that configures the parser, the Chevrotain grammar factory and language metadata. In contrast the main module is handwritten and imports
Langium's default core and shared fragments, registers any custom services such as validators, scope providers and caches, invokes \code{inject} to assemble the full
service container and exposes a \code{createServices} factory.

Although the core logic resides in `/src`, files of interest at the project root include the standard npm `package.json` and two Langium-specific configs:
\code{language-configuration.json}, which governs editor behaviors such as comment delimiters, bracket pairs and auto-closing rules, and
\code{langium-config.json}, which declares settings like the language ID, grammar entry point, target file extensions and maximum lookahead.
In particular, the `syntaxes` entry drives the generation of a `.monarch.ts` module (for Monaco-based editors) and a `.tmLanguage.json` file (for TextMate-based engines).
The TextMate grammar maps regex patterns to semantic scopes that underpin VS Code's coloring, bracket
matching and code folding, and is the one applied for the extension. Customizing these definition files allows developers to implement nuanced highlighting
- such as semantic theming and advanced pattern recognition - that cannot be exhaustively defined by grammar rules alone \cite{EDKarlsson2022Discussion604}.

Notably, the scaffolding tools also inclues further fil;es which are not needed if the sole focus is on developing the extension and language related functionsalty, and icnldued for a web demo ver..
\section{Language Server Protocol}
control flow of lsp from file to lsp to editor

@@
